# Worked normalization examples

We will demonstrate exploratory normalization steps on the expression datasets.

## Setup: download supplementary files

## load Some data


```{r get_data, include=FALSE}

knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE
)


library(GEOquery)
library(knitr)
library(dplyr)
library(tibble)
if(!require(tidyverse)){
  install.packages("tidyverse",dependencies = TRUE)
}
library(tidyverse)

gse <- "GSE233947"

source("./supp_functions.R")
fetch_geo_supp(gse = gse)

path <- file.path("data", gse)
files <- list.files(path, pattern = "\\.txt.gz$|\\.tsv.gz$|\\.csv.gz$", 
                    full.names = TRUE, recursive = TRUE)


```

Raw table preview

```{r preview_raw}
library(readr)

safe_read <- function(file) {
  # First attempt: read as TSV
  df <- tryCatch(
    readr::read_tsv(file, show_col_types = FALSE),
    error = function(e) NULL   # catch fatal errors
  )
  
  # If read_tsv failed entirely:
  if (is.null(df)) {
    message("TSV read failed — reading as space-delimited file instead.")
    return(readr::read_table(file, show_col_types = FALSE))
  }
  
  # If read_tsv returned but with parsing issues:
  probs <- problems(df)
  if (nrow(probs) > 0) {
    message("Parsing issues detected in TSV — reading as space-delimited file instead.")
    return(readr::read_table(file, show_col_types = FALSE))
  }
  
  # If everything was fine:
  return(df)
}

x <- safe_read(files[1])


kable_head(x[, 1:min(6, ncol(x))], 5, paste(gse,": raw table preview"))

```

## Example 1: GSE119732 

Before doing anything to this data.  What does it look like?

```{r}
plot_box(x[,2:ncol(x)])
```

```{r}
plot_density(x[,2:ncol(x)])
```


## Convert the raw counts to Counts per million (CPM)

```{r}
library(edgeR)

x_cpm <- cpm(y = x[,2:ncol(x)])

plot_box(x_cpm,main = "CPM - all genes")
```


Visualize as a density plot as well

```{r}
plot_density(x_cpm,main = "CPM - all genes, NO design")
```


There are a lot of genes that have zero expression.  That is the majority of them.  So we need to get rid of them.  

## filter out lowly expressed genes

```{r}

to_remove <- edgeR::filterByExpr(x_cpm,min.count = 3)
x_cpm_filtered <- x_cpm[to_remove,]

plot_box(x_cpm_filtered,main = "CPM filtered out lowly expressed - \nNO design matrix")
```

visualize this as density plot as well. 

```{r}
plot_density(x_cpm_filtered,main = "CPM filtered out lowly expressed - \n - NO design matrix")
```

## Incorporate a design matrix - description of the samples. 

The above was usingn all of the samples the same but this dataset has varying sample types.  I might be hard to figure it out just from column names as they are not so informative but let's just guess

```{r}
colnames(x)
```


```{r}

#design matrix - 
samples <- colnames(x)[2:ncol(x)]
patient  <- unlist(lapply(samples,FUN = function(x){unlist(strsplit(x,split = "_"))[2]}))       
celltype <- unlist(lapply(samples,FUN = function(x){unlist(strsplit(x,split = "_"))[3]}))     
sample_data <- data.frame(samples, patient, celltype)


design <- model.matrix(~ 0 + celltype,data = sample_data)
rownames(design) <- sample_data$samples
colnames(design) <- paste0("celltype", levels(factor(celltype)))
design


```

Filter use design information

```{r}
to_remove_withdesign <- edgeR::filterByExpr(x_cpm,
                                            min.count = 3,
                                            design = design)
x_cpm_filtered_withdesign <- x_cpm[to_remove_withdesign,]

colnames(x_cpm_filtered_withdesign ) <- paste(sample_data$celltype,
                                              1:nrow(sample_data),sep = "_")

plot_box(x_cpm_filtered_withdesign,
         main = "CPM filtered out lowly expressed - \n - with design matrix")
```

```{r}
plot_density(x_cpm_filtered_withdesign,main = "CPM filtered out lowly expressed - \n - with design matrix")
```



## Normalize Dataset using TMM

```{r}
library(edgeR)

dge <- DGEList(counts = x[,2:ncol(x)])
dge_filtered <- dge[filterByExpr(dge),]
dge_filtered  <- calcNormFactors(dge_filtered , method = "TMM")
norm_cpm <- cpm(dge_filtered , log = FALSE, prior.count = 1)

plot_box(norm_cpm)
plot_density(norm_cpm)
```

## Look at the distribution of our samples in 2D space

```{r}

y <- dge_filtered

plotMDS(y, top = 500, labels = colnames(y), 
       col = as.integer(y$samples$group))
legend("topright", legend = levels(y$samples$group),
       col = seq_along(levels(y$samples$group)), pch = 16, bty = "n")

```

Now incorporate the design into the process

```{r}

dge <- DGEList(counts = x[,2:ncol(x)],group = sample_data$celltype)
dge_filtered <- dge[filterByExpr(dge),]
dge_filtered  <- calcNormFactors(dge_filtered , method = "TMM")
norm_cpm <- cpm(dge_filtered , log = FALSE, prior.count = 1)

plot_box(norm_cpm)
plot_density(norm_cpm)
```


```{r}

y <- dge_filtered

plotMDS(y, top = 500, labels = sample_data$celltype,#colnames(y), 
       col = as.integer(y$samples$group))
legend("topright", legend = levels(y$samples$group),
       col = seq_along(levels(y$samples$group)), pch = 16, bty = "n")

```


Now incorporate the design into the process

```{r}

dge <- DGEList(counts = x[,2:ncol(x)],group = sample_data$patient)
dge_filtered <- dge[filterByExpr(dge),]
dge_filtered  <- calcNormFactors(dge_filtered , method = "TMM")
norm_cpm <- cpm(dge_filtered , log = FALSE, prior.count = 1)

plot_box(norm_cpm)
plot_density(norm_cpm)
```


```{r}

y <- dge_filtered

plotMDS(y, top = 500, labels = sample_data$patient,#  colnames(y), 
       col = as.integer(y$samples$group))
legend("topright", legend = levels(y$samples$group),
       col = seq_along(levels(y$samples$group)), pch = 16, bty = "n")

```


## Normalize the Dataset using RLE


```{r}
library(DESeq2)

counts <- x[,2:ncol(x)]

keep <- rowSums(counts >= 10) >= 2
counts_filtered <- counts[keep, ]

dds <- DESeqDataSetFromMatrix(countData = counts_filtered, 
                              colData = sample_data,
                              design = design)
dds <- estimateSizeFactors(dds)

norm_counts <- counts(dds, normalized = TRUE)

plot_box(norm_counts)
plot_density(norm_counts)
```

