
# Differential expression with edgeR (QLF) {#edger}

In this chapter we run **edgeR** using the recommended **quasi-likelihood** pipeline:

1. Start from **raw counts** in a `DGEList`
2. Estimate dispersions
3. Fit a QL GLM
4. Perform QL F-tests for contrasts of interest

We use the same additive design:

\[
\text{counts} \sim \text{cellline} + \text{treatment}
\]

```{r}
obj <- readRDS("data/d1_qc_objects.rds")
dge_f <- obj$dge_f
meta <- obj$meta
mappings <- obj$mappings

design <- obj$design

# Contrasts
contr <- makeContrasts(
  `3CTG_vs_NT` = treatment3CTG - treatmentNT,
  `20CTG_vs_NT` = treatment20CTG - treatmentNT,
  `20CTG_vs_3CTG` = treatment20CTG - treatment3CTG,
  levels = design
)

# Estimate dispersion and fit QL model
edge <- edgeR::estimateDisp(dge_f, design)
fit <- edgeR::glmQLFit(edge, design)
```

## Plot Biological coefficient of variance plot

```{r}
plotBCV(edge)
```

## Test: 3CTG vs NT and 20CTG vs NT

With this design, the coefficients `treatment3CTG` and `treatment20CTG` correspond to contrasts vs NT.

```{r}
qlf_3_vs_nt <- edgeR::glmQLFTest(fit, contrast = contr[, "3CTG_vs_NT"])
qlf_20_vs_nt <- edgeR::glmQLFTest(fit, contrast = contr[, "20CTG_vs_NT"])

edgeR::topTags(qlf_20_vs_nt)
```

## Test: 20CTG vs 3CTG

```{r}

qlf_20_vs_3 <- edgeR::glmQLFTest(fit, contrast = contr[, "20CTG_vs_3CTG"])

edgeR::topTags(qlf_20_vs_3)
```

## Extract full results

```{r}
tab_3_vs_nt <- edgeR::topTags(qlf_3_vs_nt, n = Inf)$table %>% rownames_to_column("ensembl")
tab_20_vs_nt <- edgeR::topTags(qlf_20_vs_nt, n = Inf)$table %>% rownames_to_column("ensembl")
tab_20_vs_3 <- edgeR::topTags(qlf_20_vs_3, n = Inf)$table %>% rownames_to_column("ensembl")

head(tab_20_vs_nt)
```

## Volcano plot of EdgeR results

```{r}
volcano_df <- tab_20_vs_nt %>%
  mutate(sig = FDR < 0.05)

ggplot(volcano_df, aes(x = logFC, y = -log10(PValue), color = sig)) +
  geom_point(alpha = 0.6) +
  theme_minimal() +
  labs(
    title = "EdgeR: 20CTG vs NT",
    x = "log2 fold-change",
    y = "-log10(p-value)"
  )
```


```{r}


tab <- topTags(qlf_20_vs_nt, n = Inf)$table %>%
  rownames_to_column("gene") %>%
  mutate(
    neglog10P = -log10(PValue),
    direction = case_when(
      FDR < 0.05 & logFC > 0 ~ "Up",
      FDR < 0.05 & logFC < 0 ~ "Down",
      TRUE ~ "Not sig"
    )
  )

ggplot(tab, aes(x = logFC, y = neglog10P)) +
  geom_point(aes(color = direction), alpha = 0.7, size = 1.2) +
  scale_color_manual(values = c(
    "Up" = "red",
    "Down" = "blue",
    "Not sig" = "grey70"
  )) +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "steelblue") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "steelblue") +
  theme_minimal() +
  labs(
    title = "edgeR volcano: 20CTG vs NT",
    x = "log2 fold change (logFC)",
    y = "-log10(p-value)",
    color = NULL
  )


```

## Visualize the top hits as a heatmap

 first map the ensembl ids to gene ids

```{r}

# Keep only the columns we need and ensure they are character
map_df <- mappings %>%
  transmute(
    ensembl = as.character(ensembl_gene_id),
    hgnc_symbol = as.character(hgnc_symbol)
  ) %>%
  distinct(ensembl, .keep_all = TRUE)

# Join mapping onto the edgeR results table (tab_20_vs_nt)
tab_20_vs_nt_annot <- tab_20_vs_nt %>%
  mutate(ensembl = as.character(ensembl)) %>%
  left_join(map_df, by = "ensembl") %>%
  mutate(
    # Use HGNC when available, otherwise fall back to Ensembl
    display_name = ifelse(!is.na(hgnc_symbol) & hgnc_symbol != "",
                          hgnc_symbol,
                          ensembl)
  )

# Quick check: how many mapped?
mean(!is.na(tab_20_vs_nt_annot$hgnc_symbol))
head(tab_20_vs_nt_annot[, c("ensembl","hgnc_symbol","display_name","logFC","FDR")])

```



## Select top hits from `tab_20_vs_nt_annot`

Pick either **top N by FDR** (good default for teaching) or **all genes with FDR < 0.05**.

### top N genes by FDR

```{r edger-heatmap-top-hits}
top_n <- 50  # change to 25/100 as desired

top_hits_tbl <- tab_20_vs_nt_annot %>%
  arrange(FDR, PValue) %>%
  slice_head(n = top_n)

# Ensembl IDs for subsetting the matrix
top_hits_ensembl <- top_hits_tbl$ensembl

# Row labels we want to show on the heatmap
top_hits_labels <- top_hits_tbl$display_name

head(top_hits_tbl[, c("ensembl","display_name","logFC","FDR")])
```

## Make the expression matrix for the heatmap (logCPM + row z-score)

We’ll compute **logCPM** from counts (recommended for visualization), then **z-score per gene**.

```{r edger-heatmap-matrix}
# logCPM from counts for visualization
# logCPM for visualization
logcpm <- edgeR::cpm(counts, log = TRUE, prior.count = 1)

# Keep only genes present in the expression matrix
keep <- top_hits_ensembl %in% rownames(logcpm)

hm_mat <- logcpm[top_hits_ensembl[keep], , drop = FALSE]

# Create labels aligned to hm_mat rows
hm_labels <- top_hits_labels[keep]

# IMPORTANT: avoid duplicate rownames (HGNC symbols can repeat)
# Make labels unique while preserving readability
rownames(hm_mat) <- make.unique(hm_labels)

# Z-score per gene across samples for pattern visibility
hm_z <- t(scale(t(hm_mat)))

dim(hm_z)
```

---

## Sample annotation (treatment + cellline) using Set3

```{r edger-heatmap-annotation}

meta$cellline <- as.factor(meta$cellline)
meta$treatment <- as.factor(meta$treatment)

# Set3 palettes (qualitative)
treat_levels <- levels(meta$treatment)
cell_levels  <- levels(meta$cellline)

treat_cols <- setNames(
  brewer.pal(max(3, length(treat_levels)), "Set3")[seq_along(treat_levels)],
  treat_levels
)

cell_cols <- setNames(
  brewer.pal(max(3, length(cell_levels)), "Set2")[seq_along(cell_levels)],
  cell_levels
)

ha <- HeatmapAnnotation(
  treatment = meta$treatment,
  cellline  = meta$cellline,
  col = list(
    treatment = treat_cols,
    cellline  = cell_cols
  ),
  annotation_legend_param = list(
    treatment = list(title = "Treatment"),
    cellline  = list(title = "Cell line")
  )
)

```
---

## 5) Draw the ComplexHeatmap heatmap

This produces the actual figure. Since we z-scored the rows, use a diverging palette centered at 0.

```{r edger-heatmap-plot, fig.width=8, fig.height=8}
col_fun <- circlize::colorRamp2(c(-2, 0, 2), c("blue", "white", "red"))

Heatmap(
  hm_z,
  col = col_fun,
  top_annotation = ha,
  show_row_names = FALSE,
  show_column_names = FALSE,
  cluster_rows = TRUE,
  cluster_columns = TRUE,
  row_title = paste0("Top ", nrow(hm_z), " genes (edgeR: 20CTG vs NT)"),
  column_title = "Samples"
)
```

Add labels of the gene names - only include 25 of the top hit for easier visibility.  

Is this the correct way to add the top 25?

```{r}
Heatmap(
  hm_z[1:25,],
  col = col_fun,
  top_annotation = ha,
  show_row_names = TRUE,         # <-- now shows HGNC symbols
  row_names_gp = grid::gpar(fontsize = 8),
  show_column_names = FALSE,
  cluster_rows = TRUE,
  cluster_columns = TRUE,
  row_title = paste0("Top ", nrow(hm_z), " genes (edgeR: 20CTG vs NT)"),
  column_title = "Samples"
)
```


## Save edgeR results

```{r}
saveRDS(
  list(
    meta = meta,
    counts_mapped_filtered = counts_mapped_filtered,
    fit = fit,
    qlf_3_vs_nt = qlf_3_vs_nt,
    qlf_20_vs_nt = qlf_20_vs_nt,
    qlf_20_vs_3 = qlf_20_vs_3,
    tab_3_vs_nt = tab_3_vs_nt,
    tab_20_vs_nt = tab_20_vs_nt,
    tab_20_vs_3 = tab_20_vs_3
  ),
  file = "data/d1_edger_results.rds"
)
```
