---
title: "Biomart step by step"
params:
  ensembl_version: 115
---

# biomaRt step-by-step

This follows the following workflow: **choose a mart → choose a dataset → discover filters/attributes → query with getBM() → merge back into your matrix**.

```{r}
library(biomaRt)
library(tibble)

connect_ensembl <- function(version = NULL) {
  if (!is.null(version)) {
    pinned <- tryCatch(
      useEnsembl(biomart = "ensembl", version = version),
      error = function(e) NULL
    )
    if (!is.null(pinned)) {
      return(pinned)
    }
  }

  mirrors <- c("www", "useast", "uswest", "asia")
  for (m in mirrors) {
    mart <- tryCatch(
      useEnsembl(biomart = "ensembl", mirror = m),
      error = function(e) NULL
    )
    if (!is.null(mart)) {
      return(mart)
    }
  }

  stop("Could not connect to Ensembl using pinned version or mirrors.")
}

safe_getBM <- function(attributes, filters, values, mart) {
  tryCatch(
    getBM(
      attributes = attributes,
      filters = filters,
      values = values,
      mart = mart
    ),
    error = function(e) {
      message("BioMart query unavailable in this run; using fallback with empty HGNC symbols.")
      tibble::tibble(
        ensembl_gene_id = as.character(values),
        hgnc_symbol = NA_character_
      )
    }
  )
}
```


## 1. List marts

```{r}
listMarts()
```

## 2. (Optional) pin an Ensembl archive version

```{r}
listEnsemblArchives()[1:10,]

# Example: pin to Ensembl 114
ensembl <- connect_ensembl(version = params$ensembl_version)
```

If you do not pin versions:

```{r}
ensembl <- connect_ensembl()
```

## 3. List datasets and select human

```{r}
datasets <- listDatasets(ensembl)

# filter for human
human <- datasets[grep(datasets$dataset, pattern = "hsapiens"), ]
human

ensembl <- useDataset("hsapiens_gene_ensembl", mart = ensembl)
```

## 4. Identify the correct filter

Filters are your **input ID type**.

```{r}
all_filters <- listFilters(ensembl)
dim(all_filters)

# search for Ensembl gene filters
all_filters[grep(all_filters$name, pattern = "ensembl_gene"), ]
```

For Ensembl gene IDs, use filter: `ensembl_gene_id`.

## 5. Identify the correct attributes

Attributes are the **output columns** you want.

```{r}
all_attr <- listAttributes(ensembl)
all_attr[1:10,]

# search for HGNC
searchAttributes(ensembl, "hgnc")
```

Common attributes:

- `ensembl_gene_id`
- `hgnc_symbol`

## 6. Strip version suffixes (if present)

```{r}
strip_ensembl_version <- function(x) sub("\\..*$", "", x)
ids <- c("ENSG00000141510.17", "ENSG00000157764.2")
ids_clean <- strip_ensembl_version(ids)
ids_clean
```

## 7. Run the query with `getBM()`

```{r}
map <- safe_getBM(
  attributes = c("ensembl_gene_id", "hgnc_symbol"),
  filters = "ensembl_gene_id",
  values = ids_clean,
  mart = ensembl
)
map
```

## 8. Cache the mapping (recommended)

```{r}
cache_file <- "id_conversion.rds"

if (file.exists(cache_file)) {
  map <- readRDS(cache_file)
} else {
  map <- safe_getBM(
    attributes = c("ensembl_gene_id", "hgnc_symbol"),
    filters = "ensembl_gene_id",
    values = ids_clean,
    mart = ensembl
  )
  saveRDS(map, cache_file)
}
```

## 9. Merge back into your data

```{r}
library(dplyr)

df <- tibble(ensembl_gene_id = ids_clean, value = c(1, 2))
df_annot <- left_join(df, map, by = "ensembl_gene_id")
df_annot
```


