
---
title: "Differential Expression Tutorial"
author: "BCB420 / Ruth Isserlin"
site: bookdown::bookdown_site
output:
  bookdown::html_book:
    toc: true
    toc_depth: 3
    number_sections: true
    df_print: paged
  bookdown::pdf_book:
    toc: true
    number_sections: true
documentclass: book
bibliography: [references.bib]
link-citations: true
---

# Overview

This bookdown tutorial demonstrates differential expression analysis on **GSE233947**, which provides a **FeatureCounts raw-count matrix** as a GEO supplementary file (`GSE233947_FeatureCounts_V31genes_RawCounts_ENSG.tsv.gz`).

We implement three DE workflows in clearly separated chapters:

- **limma-voom** (Chapter \@ref(limma-voom))
- **edgeR (quasi-likelihood)** (Chapter \@ref(edger))
- **DESeq2 ** (Chapter \@ref(deseq2))

In addition, it also demonstrates the analysis of differential expression analysis on a  messy dataset  **GSE240829**, which provides multiple files on GEO, two are TPM and one is labelled as salmon feature counts so for argument's sake I am going to assume that it is salmon estimated counts (as opposed to TPM, like the other files.).

- **limma-voom** (Chapter \@ref(limma-voom_dataset2))
- **edgeR (quasi-likelihood)** --> well we try quasi-likelihood (Chapter \@ref(edger_dataset2))
- **DESeq2 ** (Chapter \@ref(deseq2_dataset2))


```{r setup, include=FALSE}
# Global chunk defaults
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)

set.seed(1)
```

## Packages

```{r}
# Install once if needed:
if(!require(tidyverse)){
  install.packages(c("tidyverse"))
}
library(GEOquery)
library(edgeR)
library(limma)
library(tidyverse)
library(ComplexHeatmap)
library(RColorBrewer)
library(circlize)
```

## Reproducibility

- Run this project in RStudio.
- Click **Build** → **Build Book** (or run `bookdown::render_book("index.Rmd")`).

