# Exercises
These are intended to be done **after** completing the worked examples.

```{r setup_exercises, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(fig.width = 10, fig.height = 5)

library(GEOquery)
library(edgeR)
library(dplyr)
library(tidyr)
library(ggplot2)
library(readr)
library(tibble)
library(stringr)

source("./supp_functions.R")

safe_read <- function(file) {
  score_expression_table <- function(df) {
    if (is.null(df) || ncol(df) < 2 || nrow(df) == 0) return(-Inf)

    sample_cols <- df[, -1, drop = FALSE]
    probe_n <- min(200, nrow(sample_cols))
    if (probe_n == 0) return(-Inf)

    numeric_ratio <- vapply(sample_cols, function(x) {
      vals <- readr::parse_number(as.character(utils::head(x, probe_n)))
      mean(!is.na(vals))
    }, numeric(1))

    good_numeric_cols <- sum(numeric_ratio > 0.9)
    # Favor parses that produce many numeric sample columns (typical count matrix shape).
    good_numeric_cols * 1000 + ncol(df)
  }

  parsers <- list(
    tsv = function() read_tsv(file, show_col_types = FALSE),
    tab = function() read_delim(file, delim = "\t", show_col_types = FALSE, trim_ws = TRUE),
    ws = function() read_table(file, show_col_types = FALSE),
    csv = function() read_csv(file, show_col_types = FALSE)
  )

  parsed <- lapply(parsers, function(fn) tryCatch(fn(), error = function(e) NULL))
  scores <- vapply(parsed, score_expression_table, numeric(1))

  if (all(!is.finite(scores))) {
    stop("Could not parse expression file into gene-by-sample table: ", basename(file))
  }

  parsed[[which.max(scores)]]
}

find_expression_file <- function(gse) {
  fetch_geo_supp(gse = gse)
  path <- file.path("data", gse)
  files <- list.files(
    path,
    pattern = "\\.(txt|tsv|csv)(\\.gz)?$",
    full.names = TRUE,
    recursive = TRUE
  )

  if (length(files) == 0) stop("No expression-like supplementary files found for ", gse)

  # Prefer larger files because they are usually the counts matrix.
  files[order(file.info(files)$size, decreasing = TRUE)][1]
}

to_count_matrix <- function(df) {
  ids <- as.character(df[[1]])
  mat <- df[, -1, drop = FALSE]

  if (ncol(mat) == 0) {
    stop("No sample columns detected after removing ID column. Check delimiter/parsing.")
  }

  if (is.null(colnames(mat)) || length(colnames(mat)) == 0) {
    colnames(mat) <- paste0("sample_", seq_len(ncol(mat)))
  }
  mat[] <- lapply(mat, function(x) readr::parse_number(as.character(x)))
  mat <- as.matrix(mat)
  if (ncol(mat) > 0 && (is.null(colnames(mat)) || anyNA(colnames(mat)) || any(colnames(mat) == ""))) {
    colnames(mat) <- paste0("sample_", seq_len(ncol(mat)))
  }
  rownames(mat) <- ids

  keep_rows <- rowSums(!is.na(mat)) > 0
  mat <- mat[keep_rows, , drop = FALSE]
  mat[is.na(mat)] <- 0
  mode(mat) <- "numeric"
  mat
}

extract_model_variables <- function(meta) {
  candidates <- names(meta)[vapply(meta, function(x) is.character(x) || is.factor(x), logical(1))]
  informative <- candidates[vapply(meta[candidates], function(x) {
    nx <- length(unique(x))
    nx > 1 && nx < nrow(meta)
  }, logical(1))]

  informative[!informative %in% c("geo_accession", "title")]
}

get_geo_metadata <- function(gse, sample_names) {
  sample_names <- as.character(sample_names)
  sample_names <- sample_names[!is.na(sample_names) & sample_names != ""]

  if (length(sample_names) == 0) {
    return(tibble(sample = character()))
  }

  gse_obj <- GEOquery::getGEO(gse, GSEMatrix = TRUE, getGPL = FALSE)
  eset_list <- if (is.list(gse_obj)) gse_obj else list(gse_obj)

  pick_id_column <- function(pd) {
    id_candidates <- intersect(c("geo_accession", "title", "source_name_ch1", "description"), names(pd))
    if (length(id_candidates) == 0) {
      return(list(col = NA_character_, score = 0))
    }

    scores <- vapply(id_candidates, function(col) {
      vals <- as.character(pd[[col]])
      sum(vals %in% sample_names, na.rm = TRUE)
    }, numeric(1))

    best_col <- id_candidates[which.max(scores)]
    list(col = best_col, score = max(scores))
  }

  picks <- lapply(eset_list, function(eset) pick_id_column(pData(eset)))
  pick_scores <- vapply(picks, function(x) x$score, numeric(1))
  best_idx <- which.max(pick_scores)

  if (length(best_idx) == 0 || pick_scores[best_idx] == 0) {
    return(tibble(sample = sample_names))
  }

  pd <- pData(eset_list[[best_idx]])
  id_col <- picks[[best_idx]]$col

  meta <- tibble::as_tibble(pd) %>%
    mutate(sample = .data[[id_col]]) %>%
    filter(!is.na(sample), sample != "") %>%
    distinct(sample, .keep_all = TRUE) %>%
    select(sample, everything())

  matched <- tibble(sample = sample_names) %>%
    left_join(meta, by = "sample")

  matched
}

plot_box_colored <- function(mat, meta, color_var, main = "") {
  df_long <- as.data.frame(mat) %>%
    mutate(gene = rownames(mat)) %>%
    pivot_longer(-gene, names_to = "sample", values_to = "value") %>%
    mutate(value = log2(value + 1)) %>%
    left_join(meta %>% select(sample, !!sym(color_var)), by = "sample")

  ggplot(df_long, aes(x = sample, y = value, fill = .data[[color_var]])) +
    geom_boxplot(outlier.size = 0.15) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
    labs(title = main, x = "Sample", y = "log2(counts+1)", fill = color_var)
}

plot_density_colored <- function(mat, meta, color_var, main = "") {
  df_long <- as.data.frame(mat) %>%
    mutate(gene = rownames(mat)) %>%
    pivot_longer(-gene, names_to = "sample", values_to = "value") %>%
    mutate(value = log2(value + 1)) %>%
    left_join(meta %>% select(sample, !!sym(color_var)), by = "sample")

  ggplot(df_long, aes(x = value, colour = .data[[color_var]], group = sample)) +
    geom_density(alpha = 0.8) +
    theme_bw() +
    labs(title = main, x = "log2(counts+1)", y = "Density", colour = color_var)
}

filter_low_expression <- function(mat, meta = NULL, primary_var = NULL) {
  if (nrow(mat) == 0 || ncol(mat) == 0) {
    return(mat)
  }

  if (!is.null(meta) && !is.null(primary_var) && primary_var %in% names(meta)) {
    dge <- DGEList(counts = mat, group = factor(meta[[primary_var]]))
    keep <- filterByExpr(dge)
  } else {
    dge <- DGEList(counts = mat)
    keep <- filterByExpr(dge)
  }

  mat[keep, , drop = FALSE]
}

run_distribution_exercise <- function(gse) {
  expr_file <- find_expression_file(gse)
  raw_tbl <- safe_read(expr_file)
  mat <- to_count_matrix(raw_tbl)
  meta <- get_geo_metadata(gse, colnames(mat))
  model_vars <- extract_model_variables(meta)

  list(
    gse = gse,
    expr_file = expr_file,
    raw_tbl = raw_tbl,
    mat = mat,
    meta = meta,
    model_vars = model_vars
  )
}
```

## Exercise 1 — https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE119732

Using **GSE119732**,

1. Map the distribution of the samples as a box plot and a density plot using only the raw data.
2. Add colours to the plot separating the different variables in the model.
3. Filter out lowly expressed genes and re-plot the distributions.

```{r ex1_prepare}
ex1 <- run_distribution_exercise("GSE119732")

tibble(
  GEO = ex1$gse,
  expression_file = basename(ex1$expr_file),
  genes = nrow(ex1$mat),
  samples = ncol(ex1$mat)
) %>% knitr::kable(caption = "GSE119732 data summary")
```

### 1) Raw-data distribution plots

```{r ex1_raw_plots}
plot_box(ex1$mat, main = "GSE119732 raw counts (all genes)")
plot_density(ex1$mat, main = "GSE119732 raw counts (all genes)")
```

### 2) Colour by model variables

```{r ex1_model_vars}
knitr::kable(
  tibble(model_variable = ex1$model_vars),
  caption = "Detected model variables for GSE119732"
)
```

```{r ex1_colored_plots, fig.height=10}
if (length(ex1$model_vars) == 0) {
  message("No informative model variables detected in GEO metadata.")
} else {
  vars_to_plot <- ex1$model_vars[1:min(2, length(ex1$model_vars))]

  for (v in vars_to_plot) {
    print(plot_box_colored(ex1$mat, ex1$meta, v, main = paste("GSE119732 boxplot colored by", v)))
    print(plot_density_colored(ex1$mat, ex1$meta, v, main = paste("GSE119732 density colored by", v)))
  }
}
```

### 3) Filter low-expression genes and re-plot

```{r ex1_filtered}
ex1_primary_var <- if (length(ex1$model_vars) > 0) ex1$model_vars[1] else NULL
ex1_filtered <- filter_low_expression(ex1$mat, ex1$meta, ex1_primary_var)

tibble(
  stage = c("Before filtering", "After filtering"),
  genes = c(nrow(ex1$mat), nrow(ex1_filtered))
) %>% knitr::kable(caption = "GSE119732 gene counts before/after filtering")

plot_box(ex1_filtered, main = "GSE119732 filtered counts")
plot_density(ex1_filtered, main = "GSE119732 filtered counts")
```

## Exercise 2 — https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE122380

Using **GSE122380**, confirm whether the ID column contains Ensembl IDs with version suffixes.

1. Map the distribution of the samples as a box plot and a density plot using only the raw data.
2. Add colours to the plot separating the different variables in the model.
3. Filter out lowly expressed genes and re-plot the distributions.

```{r ex2_prepare}
ex2 <- run_distribution_exercise("GSE122380")

id_col <- names(ex2$raw_tbl)[1]
first_ids <- as.character(head(ex2$raw_tbl[[id_col]], 20))
has_version_suffix <- grepl("^ENS[A-Z]*G[0-9]+\\.[0-9]+$", first_ids)

tibble(
  GEO = ex2$gse,
  expression_file = basename(ex2$expr_file),
  genes = nrow(ex2$mat),
  samples = ncol(ex2$mat)
) %>% knitr::kable(caption = "GSE122380 data summary")

knitr::kable(
  tibble(first_20_ids = first_ids, has_version_suffix = has_version_suffix),
  caption = "GSE122380 ID check (first 20 IDs)"
)

sprintf(
  "IDs with Ensembl version suffixes in first 20 rows: %s/20",
  sum(has_version_suffix)
)
```

### 1) Raw-data distribution plots

```{r ex2_raw_plots}
plot_box(ex2$mat, main = "GSE122380 raw counts (all genes)")
plot_density(ex2$mat, main = "GSE122380 raw counts (all genes)")
```

### 2) Colour by model variables

```{r ex2_model_vars}
knitr::kable(
  tibble(model_variable = ex2$model_vars),
  caption = "Detected model variables for GSE122380"
)
```

```{r ex2_colored_plots, fig.height=10}
if (length(ex2$model_vars) == 0) {
  message("No informative model variables detected in GEO metadata.")
} else {
  vars_to_plot <- ex2$model_vars[1:min(2, length(ex2$model_vars))]

  for (v in vars_to_plot) {
    print(plot_box_colored(ex2$mat, ex2$meta, v, main = paste("GSE122380 boxplot colored by", v)))
    print(plot_density_colored(ex2$mat, ex2$meta, v, main = paste("GSE122380 density colored by", v)))
  }
}
```

### 3) Filter low-expression genes and re-plot

```{r ex2_filtered}
ex2_primary_var <- if (length(ex2$model_vars) > 0) ex2$model_vars[1] else NULL
ex2_filtered <- filter_low_expression(ex2$mat, ex2$meta, ex2_primary_var)

tibble(
  stage = c("Before filtering", "After filtering"),
  genes = c(nrow(ex2$mat), nrow(ex2_filtered))
) %>% knitr::kable(caption = "GSE122380 gene counts before/after filtering")

plot_box(ex2_filtered, main = "GSE122380 filtered counts")
plot_density(ex2_filtered, main = "GSE122380 filtered counts")
```

## Exercise 3

Can you use the worked example to process the above two GEO records? How?

Yes. The same worked-example recipe is reusable for both datasets:

1. Download GEO supplementary files.
2. Read the expression matrix safely across TSV/CSV formats.
3. Build a gene-by-sample numeric count matrix.
4. Visualize raw distributions (boxplot + density).
5. Add metadata/model-aware coloring from GEO sample annotations.
6. Filter lowly expressed genes with `edgeR::filterByExpr`.
7. Re-plot distributions after filtering to verify improved comparability.

The helper functions in this chapter (`run_distribution_exercise`, `plot_box_colored`, `plot_density_colored`, `filter_low_expression`) are the generalized implementation of that process.